maze-dataset: Maze Generation with Algorithmic Variety and Representational Flexibility

### 2025-07-30 docs/paper/paper.tex

  \begin{itemize}
  \tightlist
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/rasterized.html\#RasterizedMazeDataset}{\texttt{RasterizedMazeDataset}},
    utilizing
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.as_pixels}{\texttt{as\_pixels()}},
    which can exactly mimic the outputs provided in
    \texttt{easy-to-hard-data}
    (\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
    Bansal, et al., 2021}) and can be configured to be similar to the
    outputs of Németh (\protect\hyperlink{ref-gh_Nemeth_2019}{2019})
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.as_ascii}{\texttt{as\_ascii()}}
    provides a format similar to
    (\protect\hyperlink{ref-gh-oppenheimj2018maze}{Oppenheim, 2018};
    \protect\hyperlink{ref-eval-gpt-visual}{Singla, 2023})
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset/plotting.html\#MazePlot}{\texttt{MazePlot}}
    provides a feature‑rich plotting utility with support for multiple
    paths, heatmaps over positions, and more. This is similar to the
    outputs of (\protect\hyperlink{ref-mazegenerator-net}{Alance AB,
    2019}; \protect\hyperlink{ref-gh_Ehsan_2022}{Ehsan, 2022};
    \protect\hyperlink{ref-mathematica-maze}{Guo et al., 2011};
    \protect\hyperlink{ref-mdl-suite}{Nag, 2020})
  \end{itemize}
\item
  The text format provided by
  \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDataset.as_tokens}{\texttt{SolvedMaze(...).as\_tokens()}}
  is similar to that of (\protect\hyperlink{ref-eval-LLM-graphs}{Liu \&
  Wu, 2023}) but with many more options, detailed
  \secref{sec:tokenized-output-formats}.
\item
  Preserving metadata about the generation algorithm with the dataset
  itself is essential for studying the effects of distributional shifts.
  Our package efficiently stores the dataset along with its metadata in
  a single human-readable file (\protect\hyperlink{ref-zanj}{M.
  Ivanitskiy, n.d.}). As far as we are aware, no existing packages do
  this reliably.
\item
  Storing mazes as images or adjacency matrices is not only difficult to
  work with, but also inefficient. We use a highly efficient method
  detailed in \secref{sec:implementation}.
\item
  Our package is easily installable with source code freely available.
  It is extensively tested, type hinted, benchmarked, and documented.
  Many other maze generation packages lack this level of rigor and
  scope, and some (\protect\hyperlink{ref-ayaz2008maze}{Ayaz et al.,
  2008}) appear to simply no longer be accessible.
\end{itemize}

@@

Datasets of mazes are created from a
\docslinkcodemain{MazeDatasetConfig} configuration object, which allows
specifying the number of mazes, their size, the generation algorithm,
and various parameters for the generation algorithm. Datasets can also
be filtered after generation to satisfy certain properties. Custom
filters can be specified, and some filters are included in
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/filters.html\#MazeDatasetFilters}{\texttt{MazeDatasetFilters}}

@@

Autoregressive transformer models can be quite sensitive to the exact
format of input data, and may even use delimiter tokens to perform
reasoning steps (\protect\hyperlink{ref-pfau2024dotbydot}{Pfau et al.,
2024}; \protect\hyperlink{ref-spies2024causalworldmodels}{Spies et al.,
2024}). To facilitate systematic investigation of the effects of
different representations of data on text model performance, we provide
a variety of text output formats, with an example given in
\autoref{fig:token-regions}.

@@

@@

@@

@@

@@

@@

@@

>>>>

  \begin{itemize}
  \tightlist
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/rasterized.html\#RasterizedMazeDataset}{\texttt{RasterizedMazeDataset}},
    utilizing
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.as_pixels}{\texttt{as\_pixels()}},
    which can exactly mimic the outputs provided in
    \texttt{easy-to-hard-data}
    (\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
    Bansal, et al., 2021}) and can be configured to be similar to the
    outputs of Németh (\protect\hyperlink{ref-gh_Nemeth_2019}{2019})
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.as_ascii}{\texttt{as\_ascii()}}
    provides a format similar to
    (\protect\hyperlink{ref-gh-oppenheimj2018maze}{Oppenheim, 2018};
    \protect\hyperlink{ref-eval-gpt-visual}{Singla, 2023})
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset/plotting.html\#MazePlot}{\texttt{MazePlot}}
    provides a feature‑rich plotting utility with support for multiple
    paths, heatmaps over positions, and more. This is similar to the
    outputs of (\protect\hyperlink{ref-mazegenerator-net}{Alance AB,
    2019}; \protect\hyperlink{ref-gh_Ehsan_2022}{Ehsan, 2022};
    \protect\hyperlink{ref-mathematica-maze}{Guo et al., 2011};
    \protect\hyperlink{ref-mdl-suite}{Nag, 2020})
  \end{itemize}
\item
  The text format provided by
  \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDataset.as_tokens}{\texttt{SolvedMaze(...).as\_tokens()}}
  is similar to that of (\protect\hyperlink{ref-eval-LLM-graphs}{Liu \&
  Wu, 2023}) but with many more options, detailed in
  \hyperref[sec:tokenized-output-formats]{section: \textit{\nameref{sec:tokenized-output-formats}}}.
\item
  Preserving metadata about the generation algorithm with the dataset
  itself is essential for studying the effects of distributional shifts.
  Our package efficiently stores the dataset along with its metadata in
  a single human-readable file (\protect\hyperlink{ref-zanj}{M.
  Ivanitskiy, n.d.}). As far as we are aware, no existing packages do
  this reliably.
\item
  Storing mazes as images or adjacency matrices is not only difficult to
  work with, but also inefficient. We use a highly efficient method
  detailed in
  \hyperref[sec:implementation]{section: \textit{\nameref{sec:implementation}}}.
\item
  Our package is easily installable with source code freely available.
  It is extensively tested, type hinted, benchmarked, and documented.
  Many other maze generation packages lack this level of rigor and
  scope, and some (\protect\hyperlink{ref-ayaz2008maze}{Ayaz et al.,
  2008}) appear to simply no longer be accessible.
\end{itemize}

@@

Datasets of mazes are created from a
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDatasetConfig}{\texttt{MazeDatasetConfig}}
configuration object, which allows specifying the number of mazes, their
size, the generation algorithm, and various parameters for the
generation algorithm. Datasets can also be filtered after generation to
satisfy certain properties. Custom filters can be specified, and some
filters are included in
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/filters.html\#MazeDatasetFilters}{\texttt{MazeDatasetFilters}}.

@@

Autoregressive transformer models can be quite sensitive to the exact
format of input data, and may even use delimiter tokens to perform
reasoning steps (\protect\hyperlink{ref-pfau2024dotbydot}{Pfau et al.,
2024}; \protect\hyperlink{ref-spies2024causalworldmodels}{Spies et al.,
2024}). To facilitate systematic investigation of the effects of
different representations of data on text model performance, we provide
a variety of text output formats, with an example given in
\autoref{fig:token-regions}. We utilize Finite State Transducers
(\protect\hyperlink{ref-Gallant2015Transducers}{Gallant, 2015}) for
efficiently storing valid tokenizers.

@@

Our package is implemented in
Python(\protect\hyperlink{ref-python}{Rossum, 1995}), and makes use of
the extensive scientific computing ecosystem, including NumPy
(\protect\hyperlink{ref-numpy}{Harris et al., 2020}) for array
manipulation, plotting tools Waskom
(\protect\hyperlink{ref-seaborn}{2021}), Jupyter notebooks
(\protect\hyperlink{ref-jupyter}{Kluyver et al., 2016}), and PySR
(\protect\hyperlink{ref-pysr}{Cranmer, 2023}) for symbolic regression.

@@

\leavevmode\vadjust pre{\hypertarget{ref-pysr}{}}%
Cranmer, M. (2023). Interpretable machine learning for science with PySR
and SymbolicRegression. jl. \emph{arXiv Preprint arXiv:2305.01582}.
\url{https://doi.org/10.48550/arXiv.2305.01582}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Gallant2015Transducers}{}}%
Gallant, A. (2015). \emph{Index 1,600,000,000 keys with automata and
rust}. \url{https://burntsushi.net/transducers/}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-numpy}{}}%
Harris, C. R., Millman, K. J., Walt, S. J. van der, Gommers, R.,
Virtanen, P., \& al., et. (2020). Array programming with NumPy.
\emph{Nature}, \emph{585}, 357--362.
\url{https://doi.org/10.1038/s41586-020-2649-2}

\leavevmode\vadjust pre{\hypertarget{ref-matplotlib}{}}%
Hunter, J. D. (2007). Matplotlib: A 2D graphics environment.
\emph{Computing in Science and Engineering}, \emph{9}(3), 90--95.
\url{https://doi.org/10.1109/MCSE.2007.55}

@@

\leavevmode\vadjust pre{\hypertarget{ref-jupyter}{}}%
Kluyver, T., Ragan-Kelley, B., Perez, F., Granger, B., Bussonnier, M.,
Frederic, J., Kelley, K., Hamrick, J., Grout, J., Corlay, S., Ivanov,
P., Avila, D., Abdalla, S., \& Willing, C. (2016). Jupyter notebooks - a
publishing format for reproducible computational workflows.
\emph{Proceedings of the 20th International Conference on Electronic
Publishing}, 87--90. \url{https://doi.org/10.3233/978-1-61499-649-1-87}

@@

\leavevmode\vadjust pre{\hypertarget{ref-python}{}}%
Rossum, G. van. (1995). \emph{Python reference manual} (CS-R9525).
Centrum voor Wiskunde; Informatica (CWI).
\url{https://ir.cwi.nl/pub/5008/05008D.pdf}

@@

\leavevmode\vadjust pre{\hypertarget{ref-seaborn}{}}%
Waskom, M. L. (2021). {seaborn}: Statistical data visualization.
\emph{Journal of Open Source Software}, \emph{6}(60), 3021.
\url{https://doi.org/10.21105/joss.03021}
